Flexible Scientific Data Management for Plant Phenomics Research

### 2013-03-11 s4biodiv2013/PODD-S4BIODIV-Submission.tex
more cutbacks to reduce length

Increasingly however, providers of scientific datasets are focusing on enhancing
their datasets using curated scientific concepts from ontologies. For example,
scientists have used the Gene Ontology \cite{Ashburner2000} to link well known concepts to
represent common elements across genomics datasets. If these annotations are
given using recognised URIs, and not as key-value string tuples, various RDF
documents can be merged automatically based on RDF semantics to discover novel
conclusions. Initially, different providers were not able to decide on URIs to
use in Linked Data. This resulted in various third parties such as Bio2RDF
\cite{Belleau2008}
creating RDF versions of scientific datasets so that the RDF documents could be
easily merged based on common URIs. However, recently more providers have given
authoritative URIs for each object in their dataset, making it easy to integrate
data from various locations.

\subsection{Redesign of the Phenomics Ontology Driven Data repository}
As described earlier, the PODD repository relies on semantic web technologies to
manage phenomics data and metadata. Although both ontologies and mappings are
essential, in PODD it was necessary to build the system with a relaxed
ontological vocabulary to allow scientists to generate datasets before they were
certain about the semantic meaning of their work. In addition, it was necessary
for PODD to allow scientists to continue to maintain projects containing curated
scientific concepts alongside raw experimental data. The PODD repository was 
redesigned based on an evaluation of the original software \cite{Li2010} that 
found it was not able to scale sufficiently well to suit the HRPPC needs. The 
major differences to the old software is that projects are no longer the only 
supported top object type, and projects are not stored in multiple parts, as 
that approach was not able to scale as was originally hypothesised.

@@

An ontology is a common controlled vocabulary about a particular topic, in this
case plant phenomics. The PODD Plant ontology therefore becomes an ontology for
plant phenomics that extends the generally applicable PODD Science ontology with
plant phenomics specific details \cite{Li2010}. This approach allows PODD to model
multiple phenomics processes without enforcing a rigid data structure on
objects, either now, in the past, or in the future. In addition, other community
ontologies, such as the plant and gene ontologies \cite{Ashburner2000,Avraham2008}, can be used to
annotate the data. This not only supports data discovery, but provides common
reference points through which PODD managed data can be integrated with data in
other research databases.

The PODD application handles the instantiation and validation of objects based
on the versions of ontologies that are specified by the user as being relevant.
This gives PODD a unique extensibility as the domain knowledge is entirely
captured in the ontology, and the operational aspects of the system are derived
purely from the content of the ontology. In particular, PODD is designed around
the ability to support multiple versions of ontologies concurrently. Previously
published artifacts can be loaded without modification into PODD without
affecting current studies that may be using newer, incompatible, versions of the
schema ontologies.

PODD is able to manage high volume, high resolution, heterogeneous datasets
using a tree based structure, which in the PODD Science ontology is based on the
concept of a ``Project''. 
%(http://podd.plantphenomics.org.au/podd/). 
Data
references are inserted into projects using subproperties of the PODD
``contains'' predicate. When results are available they are inserted
independently and cross-referenced with the data that they were derived from
using a subproperty of the PODD ``refers to'' property. Each of the
subproperties have constraints on the types of objects that are valid as sources
and destinations, making it easy to both validate existing uses, and suggest
potential targets based on a source and a property.

The genotype of the plant alone does not determine the phenotype; it is the
combination of the organism’s genome with the environmental conditions that
determine the phenotype. PODD captures metadata on the genotype of the
organism(s) under investigation. PODD also gathers comprehensive metadata on the
environmental conditions and treatments that the biological samples are exposed
to. This structure makes PODD suitable to queries by web agents thus  linking
information contained in other data repositories, e.g genomics, transcriptomics,
metabolomics to phenomics information.

PlantScan\textsuperscript{\texttrademark} is a medium throughput high resolution
phenotyping platform, which brings together a number of imaging sensors--light 
detection and ranging, far-infrared imaging, and multi-wavelength imaging--to non-invasively measure plant growth and function using in-silico approaches.
% FIXME: TODO: [REF?] for PlantScan. 
Raw data is captured with
its contextual information (e.g. system configuration, time of acquisition,
batch number and project) and is stored in a purpose-built database as the data
is being generated. The various data streams are collated and used to produce
full 3D representation of each plant with overlaid spectral information. The
metadata collected during image acquisition are necessary inputs for the
computer vision techniques which are used to create the 3D representation of the
plant.
% FIXME: TODO: Create Figure 3
%(Figure 3)[e]. 
The 3D meshes are then automatically segmented in order to
semantically identify the different parts of the plants (Paproki et al, 2012). A
longitudinal 3D matching pipeline for plant mesh parts is then used to evaluate 
temporal changes at the whole plant and/or organ level. After initial
processing to recreate the 3D structure, the raw data for each plant is packaged into
one location using BagIt, an hierarchical file \cite{Kunze2011}. All bags belonging to the
same specific PODD project are grouped into collections on the CSIRO Data Access
Portal. These bags have a permanent address which is recorded into PODD. Any
additional processes on these digital objects (such as statistical information
on features like plant colour, volume,  growth etc. and on elements such as
petiole, stem, leaf) are also captured by PODD (Figure 1). Metadata
Each acquisition on PlantScan\textsuperscript{\texttrademark} includes metadata
(in addition to the raw data
streams), such as plant genus and species, project and experiment metadata, a
unique identifier for each image (Globally Unique Identifier), imaging angle,
environmental temperature of the imaging chamber, location of optical and colour
calibration datasets for each acquisition run, LiDAR (Light Detection and
Ranging Sensors) calibration files. The metadata associated with each
acquisition is automatically generated when setting up the configuration on the
platform. This information is paramount to validate and process the raw image
data, and for the post-processing phases. This information currently resides in
the PlantScan\textsuperscript{\texttrademark} database.

@@

For PlantScan\textsuperscript{\texttrademark}, file-level metadata
includes plant barcodes, batch numbers, and plant type, although the BagIt
specification does not mandate a particular archiving strategy, with the
focus being upon the directory structure, special files, and integrity checking.
% Bag creation and payload manifest integrity checking, tools can take advantage
% of the multi-core capabilities of modern computing hardware. 
To suit current
needs, BagIt-conforming tools \cite{Summers} \cite{LoCBagger} were assessed and where necessary,
improvements were implemented and tested to ensure that the tools were scalable with the size of the collections being created. 
% For example, the implementation of \cite{LoCBagger}
% assumed that as many threads as possible should be created for multi-threaded
% operations such as bag creation and payload manifest verification. On a HPC host
% with hundreds of processors, this can lead to runaway thread creation, so this
% needed to be addressed. Neither \cite{Summers} nor \cite{LoCBagger} permitted payload files in a bag to
% be listed. This has since been added to \cite{LoCBagger}. 
%Some functionality (not discussed
%here) present in [8] but not in \cite{Summers} was added to the latter.

@@

>>>>

Increasingly however, providers of scientific datasets are focusing on enhancing
their datasets using curated scientific concepts from ontologies. For example,
scientists have used the Gene Ontology \cite{Ashburner2000} to link well known concepts to
represent common elements across genomics datasets, while the Plant Ontology \cite{Avraham2008} 
allows the description of plant based datasets. If these annotations are
given using recognised URIs, and not as key-value string tuples, various RDF
documents can be merged automatically based on RDF semantics to discover novel
conclusions. Initially, different providers were not able to decide on URIs to
use in Linked Data. This resulted in various third parties such as Bio2RDF
\cite{Belleau2008}
creating RDF versions of scientific datasets so that the RDF documents could be
easily merged based on common URIs. However, recently more providers have given
authoritative URIs for each object in their dataset, making it easy to integrate
data from various locations.

\subsection{Redesign of the Phenomics Ontology Driven Data repository}
The PODD repository relies on semantic web technologies to
manage phenomics data and metadata. Although both ontologies and mappings are
essential, in PODD it was necessary to build the system with a relaxed
ontological vocabulary to allow scientists to generate datasets before they were
certain about the semantic meaning of their work. In addition, it was necessary
for PODD to allow scientists to continue to maintain projects containing curated
scientific concepts alongside raw experimental data. The PODD repository was 
redesigned based on an evaluation of the original software \cite{Li2010} that 
found it was not able to scale sufficiently well to suit the HRPPC needs. The 
major differences to the old software is that projects are no longer the only 
supported top object type, and projects are not stored in multiple parts, as 
that approach was not able to scale as was originally hypothesised.

@@

% An ontology is a common controlled vocabulary about a particular topic, in this
% case plant phenomics. The PODD Plant ontology therefore becomes an ontology for
% plant phenomics that extends the generally applicable PODD Science ontology with
% plant phenomics specific details \cite{Li2010}. This approach allows PODD to model
% multiple phenomics processes without enforcing a rigid data structure on
% objects, either now, in the past, or in the future. In addition, other community
% ontologies, such as the plant and gene ontologies \cite{Ashburner2000,Avraham2008}, can be used to
% annotate the data. This not only supports data discovery, but provides common
% reference points through which PODD managed data can be integrated with data in
% other research databases.

% The PODD application handles the instantiation and validation of objects based
% on the versions of ontologies that are specified by the user as being relevant.
% This gives PODD a unique extensibility as the domain knowledge is entirely
% captured in the ontology, and the operational aspects of the system are derived
% purely from the content of the ontology. In particular, PODD is designed around
% the ability to support multiple versions of ontologies concurrently. Previously
% published artifacts can be loaded without modification into PODD without
% affecting current studies that may be using newer, incompatible, versions of the
% schema ontologies.

% PODD is able to manage high volume, high resolution, heterogeneous datasets
% using a tree based structure, which in the PODD Science ontology is based on the
% concept of a ``Project''. 
% %(http://podd.plantphenomics.org.au/podd/). 
% Data
% references are inserted into projects using subproperties of the PODD
% ``contains'' predicate. When results are available they are inserted
% independently and cross-referenced with the data that they were derived from
% using a subproperty of the PODD ``refers to'' property. Each of the
% subproperties have constraints on the types of objects that are valid as sources
% and destinations, making it easy to both validate existing uses, and suggest
% potential targets based on a source and a property.

% The genotype of the plant alone does not determine the phenotype; it is the
% combination of the organism’s genome with the environmental conditions that
% determine the phenotype. PODD captures metadata on the genotype of the
% organism(s) under investigation. PODD also gathers comprehensive metadata on the
% environmental conditions and treatments that the biological samples are exposed
% to. This structure makes PODD suitable to queries by web agents thus  linking
% information contained in other data repositories, e.g genomics, transcriptomics,
% metabolomics to phenomics information.

@@

For PlantScan\textsuperscript{\texttrademark}, file-level metadata
includes plant barcodes, batch numbers, and plant type, although the BagIt
specification does not mandate a particular archiving strategy, with the
focus being upon the directory structure, special files, and integrity checking.
% Bag creation and payload manifest integrity checking, tools can take advantage
% of the multi-core capabilities of modern computing hardware. 
% To suit current
% needs, 
BagIt-conforming tools \cite{Summers} \cite{LoCBagger} were assessed and where necessary,
improvements were implemented and tested to ensure that the tools were scalable with the size of the collections being created. 
% For example, the implementation of \cite{LoCBagger}
% assumed that as many threads as possible should be created for multi-threaded
% operations such as bag creation and payload manifest verification. On a HPC host
% with hundreds of processors, this can lead to runaway thread creation, so this
% needed to be addressed. Neither \cite{Summers} nor \cite{LoCBagger} permitted payload files in a bag to
% be listed. This has since been added to \cite{LoCBagger}. 
%Some functionality (not discussed
%here) present in [8] but not in \cite{Summers} was added to the latter.

@@

PlantScan\textsuperscript{\texttrademark} is a medium throughput high resolution
phenotyping platform, which brings together a number of imaging sensors--light 
detection and ranging, far-infrared imaging, and multi-wavelength imaging--to non-invasively measure plant growth and function using in-silico approaches.
% FIXME: TODO: [REF?] for PlantScan. 
Raw data is captured with
its contextual information (e.g. system configuration, time of acquisition,
batch number and project) and is stored in a purpose-built database as the data
is being generated. The various data streams are collated and used to produce
full 3D representation of each plant with overlaid spectral information. The
metadata collected during image acquisition are necessary inputs for the
computer vision techniques which are used to create the 3D representation of the
plant.
% FIXME: TODO: Create Figure 3
%(Figure 3)[e]. 
The 3D meshes are then automatically segmented in order to
semantically identify the different parts of the plants (Paproki et al, 2012). A
longitudinal 3D matching pipeline for plant mesh parts is then used to evaluate 
temporal changes at the whole plant and/or organ level.

% After initial processing to recreate the 3D structure, the raw data for each plant is packaged into
% one location using BagIt. All bags belonging to the
% same specific PODD project are grouped into collections on the CSIRO Data Access
% Portal. These bags have a permanent address which is recorded into PODD. Any
% additional processes on these digital objects (such as statistical information
% on features like plant colour, volume,  growth etc. and on elements such as
% petiole, stem, leaf) are also captured by PODD using the properties in the 
% PODD Plant Ontology (Figure~\ref{poddontologies}).

\subsection{Metadata}
Each acquisition on PlantScan\textsuperscript{\texttrademark} includes metadata
(in addition to the raw data
streams), such as plant genus and species, project and experiment metadata, a
unique identifier for each image (Globally Unique Identifier), imaging angle,
environmental temperature of the imaging chamber, location of optical and colour
calibration datasets for each acquisition run, LiDAR (Light Detection and
Ranging Sensors) calibration files. The metadata associated with each
acquisition is automatically generated when setting up the configuration on the
platform. This information is paramount to validate and process the raw image
data, and for the post-processing phases. This information currently resides in
the PlantScan\textsuperscript{\texttrademark} database.
